\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-25 coqimpl.tex
edits

\YZ{1. move some of sec 3.2 ~> 3.1; 2. hint vaguely on how to do type checking in sec 3.1 (user interface).. metatheory part should also mention this part in sec 3.1} This is achieved by a complete compilation from a family to a module, via aggregating the list of the modules in the internal representation: the plugin will repeatedly using Coq's "Include" command as the "Module A" in \cref{fig:plugin-example1} illustrates. Now, the user can use "Module A" anywhere as they are using "Family A".
\YZ{The reader wonders why this aggregation is needed. 
Seems like the place to talk about how the programmer can use A outside Family A.}\EDJreply{Please check.}

Note that, in our example, we include two
functors without arguments---this is due to a peculiarity of Coq's
"Include": uninstantiated parameters will be automatically
instantiated by the appropriate fields in the surrounding defining
module.
For example, "b_7" requires a module of type "a_6". Inside module "A",
when "Include a_4" is done, we will have a field \mintinline{Coq}{a : nat}
inside the current surrounding "A". Then when we
\mintinline{Coq}{Include b_7} without specifying the module parameter,
Coq will try to find fields in the surrounding module (i.e., module
"A") to instantiate \mintinline{Coq}{self_A: a_6}. This is satisfied by
the \mintinline{Coq}{a : nat} that is included earlier.
The resulting compiled module is just like mundane Coq's module and 
thus maintains all the computational mechanism from Coq, fulfilling requirements in \ref{chg:software-engineering}.

@@

Simiarly, we
use a list to encode the three kinds of inheritance data in the
implementation. The inheritance data will include the
information of (1) the ``type'' of the parent family, (2) a list of
operation indicating how each field of the parent will be dealt with
(either inherited or overriden), (3) the operations indicating newly
extended fields, and (4) the ``type'' of the children family.
The surface syntax for the three kinds of inheritance
are demonstrated in the exemplar \cref*{fig:plugin-example3}. The
implementation of overriding, inheritance, and extension is achieved by
simply manipulating the internal representation of a family
correspondingly---we can swap the module in the list for overriding,
retain the module in the list for inheritance, and add new module into
the list for extension. The compilation from a family (internal
representation) to a module still acts the same.

@@

Our implementation also make sure the first attempt of defining "c" failed successfully in \cref{fig:plugin-example3}---the proof of "c" will require a concrete definition of "a",
but our compilation will type check \mintinline{Coq}{c : a = 1 := eq_refl} in (a module with) the context where we only know \mintinline{Coq}{a : nat} thus failed.

\textbf{Implemening "Final" and "Sealed Family".}
We implement "Final" by
exposing the whole definition directly into the compiled module type. Doing so will prohibit overriding. Say, the example "bop" in \cref{fig:plugin-example3}, will be prohibited to be overridden by $(-)$
subtraction doesn't have the same definition as $(+)$. We require the
users to manual decorate field as `Final' so that the plugin will proceed with this special treatment.
%\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on talking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. Don't resolve it until I add it into Figure 2.}

@@

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We support extensible inductive types in Coq mentioned in
\ref{chg:extensible-inductive-type} but we will not be aiming at the research of its formal semantic.\footnote{Because to give a satisfactory formal definition and semantic for extensible inductive type is not something this paper aims at, and obviously requires more theoretical research effort}\YZ{I do consider that your plugin supports extensible inductive types. It's just that they are implemented as a plugin, rather than in the Coq core.}\EDJreply{I rephrased it. It is more like I don't think this is really the final form of extensible inductive type that everyone is happy about}

@@

Module Type STLC_409.
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* ... and all the fields defined before
  and the abstraction for inductive type ...*)
End STLC_409.
(* Collect eliminator to Set for tm *)
Module tm_rec_425 (STLC: STLC_409).
  Definition __recursor_type_tm_rec :=
    forall P : STLC.tm -> Set,
    (forall n : id, P (STLC.tm_var n)) ->
    (forall (n : id) (i : STLC.tm),
    P i -> P (STLC.tm_abs n i)) ->
    ... -> forall i : STLC.tm, P i.
End tm_rec_425.
(* Compiled Field STLC.subst_internal.tm_var *)
Module tm_var_411 (STLC: STLC_409)
  (subst_internal: EmptySig).
  Definition tm_var :
    forall (s : id) (x : id) (t : STLC.tm), STLC.tm :=
    fun s x t ⇒ if eqb x s then t else STLC.var s.
End tm_var_411. (* ... and more *)
(* Aggregate into one subst_internal*)
Module subst_internal_410 (STLC: STLC_409). 
  Include tm_var_411 STLC. (* And other cases ... *)
End subst_internal_410.
Module v_33_34 (STLC: STLC_409).
(* Intermediate Module solely for type checking *)
  Include tm_rec_425 STLC.
  Parameter (recursor_for_type_checking : __recursor_type_tm_rec).
  Definition term_for_type_checking :=
    recursor_for_type_checking (fun _  ⇒ id → STLC.tm → STLC.tm)
      subst_internal_410.tm_var
      subst_internal_410.tm_abs ... .
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Exempler STLC Code, especially about Inductive Type}\label{fig:plugin-example2}
\end{figure}\YZ{Change this one into STLC example, to see the layout}\EDJreply{Changed. I think it only make things worse.}

@@

Let's look at the exemplar \cref{fig:plugin-example2}, the example of "subst" on STLC. When we define the extensible inductive type "tm", all the following fields (e.g. "subst_internal.tm_var") can only know "tm" is a type (of "Set") and there are at least two constructors for it, just like what we declare in ``the abstracted interface''---the compiled module type "STLC_409". We cannot export\YZ{export or put?}\EDJreply{Ok. I just figure out how my wording is confusing. put is absolutely correct but by put I want to mean ``export''. I think I use ``export'' in several places...}
the eliminator of the inductive type "tm" to "STLC_409" because doing so all of the following fields (e.g. "subst_internal.tm_var" and the corresponding "tm_var_411") will not be able to be inherited to the context where "tm"
is extended with a third constructor, \textbf{since the mere existence of the recursor will assert that there are only two constructors} in "STLC_bool", not
\textbf{extensible} at all!\YZ{I can see how the existence of the recursor alone prevents extensibility, but I don't see why it has to do with ff_24.}\EDJreply{You are right. It is not only about "ff_24".I rephrased it. Please check.}

However, we later still need the concrete eliminator on "tm" for doing exhaustiveness checking (as in \ref{chg:extensible-inductive-type}), thus we will store the corresponding info somewhere.  Here we show one exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module "tm_rec_425". This concrete eliminator will be used by our "FRecursor". For example, in \cref{fig:plugin-example2}, "FRecursor" command will feed Coq a module "v_33_34" for doing type-checking with specified handlers (e.g. "neg_handler") during compilation---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq. Note that, at this point of the example, "FRecursor" will only do type check. The compilation of "FRecursor" into a component of the compiled module only happens when closing the whole family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla) inductive type in that compiled module. Thus a "FRecursor" field is unlike other fields, even though it is exposed as a mundane function for the following fields in the programming interface.

The plugin actually realizes the "Extend FInductive" via overridding---when extending an inductive type $A$
with new constructors $c$, our plugin will generate a new (syntactic) inductive definition $A'$ with this new $c$ on the fly and feed it to Coq. But we must be careful on the exposing data in the family type (the compiled module type) for the following field. In other words, we must be careful on what interface (context) the following fields are defined based on.

Importantly, the recursor field (e.g. "subst") is not actually inheritable, because it is not compiled like any other fields---recall how other fields are compiled into reusable pieces of (parametrized) modules, a recursor field is not compiled like that. For all the children family, we have to reconstruct and override the "FRecursor" field with a ``new'' recursor using (maybe the same) handler family. This is affordable since each aggregated recursor handler doesn't need
to be rechecked---look at "STLC_bool.neg_handler", there are only  new
handlers for "tm_true" and the handlers (mundane fields) from the parent
"B.subst_internal.tm_var" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "STLC_409"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "tm_var_411"---doesn't require another type-check and is re-used during the type-checking and compilation of "STLC_bool.subst". For example, for the new "subst", when compiling family "STLC_bool" into a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla but extended and modified) inductive type in that compiled module. Note that, even though "tm_rec" and "tm" has the same name as before, they are in the newly-compiled module so they refer to totally different things.

Of course, in the implementation, our plugin will ``inherit'' the
recursor "neg" in the surface \YZ{Show this recursor, as you do above for the recursor "Definition neg := ...".}\EDJreply{Added above. Please Check.}
but what the plugin actually does is doing a second exhaustiveness
checking in children family like mentioned above, with the ``same'' motive and the ``same'' handler family.\YZ{A source of confusion in this section is that the reader often has to guess who performs the actions: is it the programmer or the Coq plugin?}\EDJreply{Good idea. Let me reorgnaize this section to ``first programming interface'', then ``plugin implementation''}\EDJreply{Please check.}

>>>>

This is achieved by a complete compilation from a family to a module,
via aggregating the list of the modules in the internal representation:
the plugin will repeatedly using Coq's "Include" command as the "Module
A" as \cref{fig:plugin-example1} illustrates.
References to family "A" outside "A" will be compiled to references to
this Coq module "A".

Note that, in our example, we include two
functors without arguments---this is due to a peculiarity of Coq's
"Include": uninstantiated parameters will be automatically
instantiated by the appropriate fields in the surrounding defining
module.
For example, "b_7" requires a module of type "a_6". Inside module "A",
when "Include a_4" is done, we will have a field \mintinline{Coq}{a : nat}
inside the current surrounding "A". Then when we
\mintinline{Coq}{Include b_7} without specifying the module parameter,
Coq will try to find fields in the surrounding module (i.e., module
"A") to instantiate \mintinline{Coq}{self_A: a_6}. This is satisfied by
the \mintinline{Coq}{a : nat} that is included earlier.
The resulting compiled module is just like mundane Coq's module and 
thus maintains all the computational mechanism from Coq, fulfilling
requirements in \ref{chg:software-engineering}.\YZ{
  Do I understand correctly that an alternative to individually
  compiling family members to modules is to compile a family as a whole,
  but this alternative would not satisfy the "incremental type checking"
  or "instant feedback" requirement?
}\YZreply{
  Looks like individually compiling family members addresses not only
  this requirement, but also the "self vs. consistency" challenge.
}

@@

Simiarly, we
use a list to encode the three kinds of inheritance data in the
implementation. The inheritance data will include the
information of (1) the ``type'' of the parent family, (2) a list of
operation indicating how each field of the parent will be dealt with
(either inherited or overriden), (3) the operations indicating newly
extended fields, and (4) the ``type'' of the children family.
The surface syntax for the three kinds of inheritance
are demonstrated in the exemplar \cref{fig:plugin-example3}. The
implementation of overriding, inheritance, and extension is achieved by
simply manipulating the internal representation of a family
correspondingly---we can swap the module in the list for overriding,
retain the module in the list for inheritance, and add new module into
the list for extension. The compilation from a family (internal
representation) to a module still acts the same.

@@

Our implementation also makes sure the first attempt of defining "c"
failed successfully in \cref{fig:plugin-example3}---proving "c"
would require a concrete definition of "a", and our plugin rejects
\mintinline{Coq}{c : a = 1 := eq_refl} in a context where we
only know \mintinline{Coq}{a : nat}.

\textbf{Implementing "Final" and "Sealed Family".}
We implement "Final" by
exposing the whole definition directly into the compiled module type.
Doing so will prohibit overriding. Say, "bop", declared "Final" in
\cref{fig:plugin-example3}, is not overridable.
We require the users to manual decorate field as `Final' so that the
plugin will proceed with this special treatment.
%\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on talking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. Don't resolve it until I add it into Figure 2.}

@@

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We support extensible inductive types in Coq mentioned in
\ref{chg:extensible-inductive-type}, but we will not be aiming at the
research of its formal semantic.\footnote{Because to give a satisfactory
formal definition and semantic for extensible inductive type is not
something this paper aims at, and obviously requires more theoretical
research effort}
%\YZ{I do consider that your plugin supports extensible inductive types. It's just that they are implemented as a plugin, rather than in the Coq core.}\EDJreply{I rephrased it. It is more like I don't think this is really the final form of extensible inductive type that everyone is happy about}

@@

Module Type STLC_409.
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* ... and all the fields defined before
  and the abstraction for inductive type ...*)
End STLC_409.
(* Collect eliminator to Set for tm *)
Module tm_rec_425 (STLC: STLC_409).
  Definition __recursor_type_tm_rec :=
    forall P : STLC.tm -> Set,
    (forall n : id, P (STLC.tm_var n)) ->
    (forall (n : id) (i : STLC.tm),
    P i -> P (STLC.tm_abs n i)) ->
    ... -> forall i : STLC.tm, P i.
End tm_rec_425.
(* Compiled Field STLC.subst_internal.tm_var *)
Module tm_var_411 (STLC: STLC_409)
  (subst_internal: EmptySig).
  Definition tm_var :
    forall (s : id) (x : id) (t : STLC.tm), STLC.tm :=
    fun s x t ⇒ if eqb x s then t else STLC.var s.
End tm_var_411. (* ... and more *)
(* Aggregate into one subst_internal*)
Module subst_internal_410 (STLC: STLC_409). 
  Include tm_var_411 STLC. (* And other cases ... *)
End subst_internal_410.
Module v_33_34 (STLC: STLC_409).
(* Intermediate Module solely for type checking *)
  Include tm_rec_425 STLC.
  Parameter (recursor_for_type_checking : __recursor_type_tm_rec).
  Definition term_for_type_checking :=
    recursor_for_type_checking (fun _  ⇒ id → STLC.tm → STLC.tm)
      subst_internal_410.tm_var
      subst_internal_410.tm_abs ... .
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Exemplar STLC Code, especially about Inductive Type}\label{fig:plugin-example2}
\end{figure}

@@

Let's look at the exemplar \cref{fig:plugin-example2}, the example of
"subst" on STLC. When we define the extensible inductive type "tm", all
the following fields (e.g., "subst_internal.tm_var") can only know "tm"
is a type (of "Set") and there are at least two constructors for it,
just like what we declare in ``the abstracted interface''---the compiled
module type "STLC_409". We cannot export
%\YZ{export or put?}\EDJreply{Ok.
%I just figure out how my wording is confusing. put is absolutely correct
%but by put I want to mean ``export''. I think I use ``export'' in
%several places...}
the eliminator of the inductive type "tm" to "STLC_409".
Otherwise all the subsequent fields (e.g. "subst_internal.tm_var" and
the corresponding "tm_var_411") would not be able to be inherited to the
context where "tm" is extended with a third constructor, \textbf{since
the mere existence of the recursor will assert that there are only two
constructors} in "STLC_bool", not \textbf{extensible} at all!

However, we later still need the concrete eliminator on "tm" for doing
exhaustiveness checking (as in \ref{chg:extensible-inductive-type}),
thus we will store the corresponding info somewhere.  Here we show one
exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module
"tm_rec_425". This concrete eliminator will be used by our "FRecursor".
For example, in \cref{fig:plugin-example2}, "FRecursor" command will
feed Coq a module "v_33_34" for doing type-checking with specified
handlers (e.g. "neg_handler") during compilation---what it does is
basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.\YZ{
  Sounds like we can still claim the plugin is doing type checking.
  It's just that the checking is done by generating some Coq code and
  then having Coq check it.
}
Note that, at this point of the example,
"FRecursor" will only do type check. The compilation of "FRecursor" into
a component of the compiled module only happens when closing the whole
family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and
the (vanilla) inductive type in that compiled module. Thus a "FRecursor"
field is unlike other fields, even though it is exposed as a mundane
function for the following fields in the programming interface.\YZ{
  Why is this special treatment needed for compiling FRecursor?
}

The plugin actually realizes the "Extend FInductive" via overriding---%
when extending an inductive type $A$ with new constructor $c$, our
plugin will generate a new (syntactic) inductive definition $A'$ with
this new $c$ on the fly and feed it to Coq. But we must be careful on
the exposing data in the family type (the compiled module type) for the
following field. In other words, we must be careful on what interface
(context) the following fields are defined based on.

Importantly, the recursor field (e.g. "subst") is not actually
inheritable, because it is not compiled like any other fields.
Recall how other fields are compiled into reusable pieces of
(parametrized) modules; a recursor field is not compiled like that.
For all the children family, we have to reconstruct and override the
"FRecursor" field with a ``new'' recursor using (maybe the same) handler
family.  This is affordable since each aggregated recursor handler
doesn't need to be rechecked---look at "STLC_bool.neg_handler", there
are only  new handlers for "tm_true" and the handlers (mundane fields)
from the parent "B.subst_internal.tm_var" have already been type-checked
and inherited (since they only rely on ``the abstracted interface''
module "STLC_409"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "tm_var_411"---doesn't require another type-check and is
re-used during the type-checking and compilation of "STLC_bool.subst".
For example, for the new "subst", when compiling family "STLC_bool" into
a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla
but extended and modified) inductive type in that compiled module. Note
that, even though "tm_rec" and "tm" has the same name as before, they
are in the newly-compiled module so they refer to totally different
things.

Of course, in the implementation, our plugin will ``inherit'' the
recursor "neg" in the surface,
but what the plugin actually does is doing a second exhaustiveness
checking in children family like mentioned above, with the ``same'' motive and the ``same'' handler family.
%\YZ{A source of confusion in this section is that the reader often has to guess who performs the actions: is it the programmer or the Coq plugin?}\EDJreply{Good idea. Let me reorgnaize this section to ``first programming interface'', then ``plugin implementation''}

### 2022-07-25 introduction.tex
edits

At the core of many linguistic solutions to the EP is \emph{inheritance}.
Inheritance is sometimes interpreted narrowly as a subclass'
inheriting methods and fields from a superclass.
But its language-theoretic essence is more general:
inheritance is a linguistic approach to incrementally
modifying \emph{recursive, record-like definitions}~\cite{cook1990inheritance}.

@@

Integrating family polymorphism into a dependent type theory with
inductive types poses significant technical challenges, however.
As we analyze, pillars of such a dependent type theory—including
inductive types, definitional equalities, and logical consistency—are
all inimical to the kind of extensibility and family polymorphism
found in existing OO language designs.
Thus, our contributions include novel design recipes for dealing with
these challenges and strong meta\-theoretical guarantees on the
underlying logic.

>>>>

At the core of many linguistic solutions to the EP is \emph{inheritance}.
Inheritance is sometimes interpreted narrowly as a subclass'
inheriting methods and instance variables from its superclass.
But the language-theoretic essence of inheritance is more general:
it is a linguistic approach to incrementally modifying
\emph{recursive, record-like definitions}~\cite{cook1990inheritance}.

@@

Integrating family polymorphism into a dependent type theory for
logical reasoning, however, poses significant technical challenges.
As we analyze, pillars of dependent type theories—including
inductive types, definitional equality, and logical consistency—are
all inimical to the kind of extensibility and family polymorphism
found in existing OO language designs.
Thus, our contributions include novel design recipes for dealing with
these challenges and strong meta\-theoretical guarantees on the
underlying logic.

### 2022-07-25 syntax.tex
edits

In other words, (1) a universe $\cU$ is a type including the \textit{codes or names} of types.
(2) a code (term) $\goodTerm{\Gamma}{T}{\cU}$ is not allowed to locate at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ is (i.e. we distinguish the code and the type). (3)
"c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of a
type. More concretely, using QIIT, the corresponding constructors will look like, (notice that the type on both sides of the equalities are always the same and thus a well-typed equality) 
\begin{align*}
  &\mintinline{Agda}{c : Ty Γ → Tm Γ U} &&\mintinline{Agda}{El : (a : Tm Γ U) → Ty } \\
  &\mintinline{Agda}{Elc : El (c T) ≡ T} && \mintinline{Agda}{cEl : c (El T) ≡ T}
\end{align*}
These two equations indicate how \textit{dynamic} should react when $El~"c"~T$ appears. Later we will see other similar $\beta$-equations but we won't display their (pseudo-)Agda formulation.

>>>>

In other words, (1) a universe $\cU$ is a type inhabited by the \textit{codes or names} of types.
(2) a code (term) $\goodTerm{\Gamma}{T}{\cU}$ cannot be used at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ can (i.e., we distinguish the code and the type).
(3) "c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of
a type. More concretely, using QIIT, the corresponding constructors will
look like, (notice that the type on both sides of the equalities are
always the same and thus a well-typed equality) 
\begin{align*}
  &\mintinline{Agda}{c : Ty Γ → Tm Γ U} &&\mintinline{Agda}{El : (a : Tm Γ U) → Ty } \\
  &\mintinline{Agda}{Elc : El (c T) ≡ T} && \mintinline{Agda}{cEl : c (El T) ≡ T}
\end{align*}
These two equations indicate how \textit{dynamic} should react when $El~"c"~T$ appears. Later we will see other similar $\beta$-equations but we won't display their (pseudo-)Agda formulation.
